``Interestingness'' in Mathematical Theory Exploration

### 2019-11-29 PhDThesis/background.tex
Thesis bits
Headingy things
Background changes
TODOs
Remove some duplicate text
A bunch of cleanup

\section{Background}

\input{background_haskell.tex}
\input{background_quickcheck.tex}
\input{background_theoryexploration.tex}
\input{background_conjecturegeneration.tex}
\input{background_machinelearning.tex}
\input{background_recurrentclustering.tex}

>>>>

\chapter{Background}

% Background should cover approaches to automated reasoning about software, and
% formal systems in general. We can highlight how the language/representation
% imposes invariants and guarantees on what we can know, and how we can extend
% those guarantees with features like purity (prevents behaviour depending on
% ambient environment/state), lazy evaluation (prevents problems with
% non-termination), sound type systems (limits the space of possible values for
% each expression), etc. and therefore how Haskell is a natural vehicle for us to
% use. Giving background on Haskell naturally leads to descendants like Agda and
% Idris, which are also interactive theorem provers; which lets us discuss
% Isabelle, Coq, HOL, etc., how they're difficult for existing reasoning
% algorithms to handle (e.g. resolution, which works well in first-order logics),
% approaches to helping

This chapter provides background on the fields we investigate, and the
techniques we apply. It includes descriptions of the proof- and
programming-languages we use, and justifications for their use in our work. A
very short summary of relevant Machine Learning topics is also given, to provide
context for the techniques we use (namely unsupervised clustering) and the
potential alternatives and future directions that are most applicable.

\input{background_haskell.tex}
\input{background_quickcheck.tex}
\input{background_theoryexploration.tex}

% FIXME: This seems to be repetitive of Theory Exploration?
% \input{background_conjecturegeneration.tex}

\input{background_machinelearning.tex}

### 2019-11-29 PhDThesis/background_machinelearning.tex
Remove some duplicate text
A bunch of cleanup

\emph{Feature extraction} is a common pre-processing step for machine learning
(ML). Rather than feeding ``raw'' data straight into our ML algorithm, we only
learn a sample of \emph{relevant} details, known as \emph{features}. This has
two benefits:

@@

 One common approach to this problem is
to represent such structures as \emph{sequences}. \emph{Recurrent neural
  networks} (RNNs) are a popular choice for processing sequences, especially
when combined with mechanisms such as \emph{long short-term memory} (LSTM) for
preserving information across long sequences \cite{hochreiter1997long}. Such
systems have been used, for example, to parse and execute computer programs
\cite{zaremba2014learning}. However, learning to parse sequences seems
inefficient considering that we already have correctly-parsed ASTs.

Whilst neural networks have been applied directly to recursive structures
\cite{goller1996learning}, including using LSTM \cite{zhu2015long}, a more
popular approach is to use \emph{kernel methods}
\cite{bakir2007predicting}. These are promising as a more principled alternative
to our current hand-crafted translation of ASTs to vectors.

\input{background_clustering.tex}

>>>>

\emph{Feature extraction} is a common pre-processing step for machine learning
(ML). Rather than feeding ``raw'' data straight into our ML algorithm, we only
learn a sample of \emph{relevant} details, known as \emph{features}. This has
a few benefits:

@@

% FIXME: Include clustering

### 2019-11-29 PhDThesis/background_theoryexploration.tex
A bunch of cleanup

\iffalse
\quickspec{} (and \hipspec{}) is also compatible with Haskell's existing testing
infrastructure, such that an invocation of \texttt{cabal test} can run these
tools alongside more traditional QA tools like \quickcheck{}, \textsc{HUnit} and
\textsc{Criterion}.

In fact, there are similarities between the way a TE system like \quickspec{} can
generalise from checking \emph{particular} properties to \emph{inventing} new
ones, and the way counterexample finders like \quickcheck{} can generalise from
testing \emph{particular} expressions to \emph{inventing} expressions to
test. One of our aims is to understand the implications of this generalisation,
the lessons that each can learn from the other's approach to term generation,
and the consequences for testing and QA in general.
\fi

% TODO
%%%%% THE BELOW SEEMS TO BE INTERESTINGNESS?

%\begin{description}
%\item{Interestingness} Various alternative interestingness criteria have been
%  proposed, which we survey in \S \ref{sec:relatedwork}. Augmenting or replacing
%  the criteria may be useful, for example to distinguish useful relationships
%  from incidental coincidences; or to prevent surprising, insightful equations
%  from being discarded because they can be simplified.
%\end{description}

% Theory exploration is similar to \emph{experimental mathematics}

% - Relation to Science
%  - Testable/falsifiable hypotheses are like evaluable terms (or, more generally, conjectures which can be decided, using a reasonable amount of resources).
% - Relation to AI tasks: exploring surroundings, etc.

% - Statistics is another area that's less straightforward than normal numerical computing, since there is subjectivity and judgement involved in the answering of questions.

% Theory formation: Alison? Others.
% Theory exploration: Buchberger, Moa in Isabelle, Koen in Haskell. Others?
% Theorem proving: Well-trodden: first-order ATP, higher-order ITP, functional programming
% Communication: Latex, Wikis, APIs, communicating with aliens

>>>>

\subsection{Related Fields}

\quickspec{} (and \hipspec{}) grew out of software engineering tools like
\quickcheck{}, and they are in fact compatible with Haskell's existing testing
infrastructure, such that an invocation of \texttt{cabal test} can run these
tools alongside more traditional QA tools like \quickcheck{}, \textsc{HUnit} and
\textsc{Criterion}.

There are similarities between the way a TE system like \quickspec{} can
generalise from checking \emph{particular} properties to \emph{generating} new
ones, and the way counterexample finders like \quickcheck{} can generalise from
testing \emph{particular} expressions to generating expressions to test.

\emph{Experimental mathematics} is related to theory exploration, in that both
seek to use empirical methods to discover high-level properties of
well-specified formal systems. The theory is itself may be a scientific model,
in which case theory exploration becomes part of the scientific method; in
particular, finding ``testable predictions'' (e.g. equations relating measurable
quantities, which can be experimentally tested) is a special case of theory
exploration's search for ``interesting conjectures''.

\emph{Theory formation} and \emph{concept formation} go beyond mere exploration;
inventing new definitions, which can give rise to subsequent properties. This
gives rise to a feedback loop: concepts are ``interesting'' if they have
``interesting'' properties, whilst properties are ``interesting'' due to the
concepts they involve. This gives rise to some deep philosophical questions
about the nature of mathematics, but is so open-ended that it can be difficult
to apply to day-to-day tasks; hence why we generally avoid invention systems
in this work.

### 2019-11-29 PhDThesis/bucketing.tex
A bunch of cleanup

>>>>

\iffalse
Our approach to scaling up \quickspec{} takes inspiration from two sources. The
first is relevance filtering, which makes expensive algorithms used in theorem
proving more practical by limiting the size of their inputs. We describe this
approach in more details in \S~\ref{sec:relevance}. Relevance filtering is a
practical tool which has existing applications in software, such as the
\emph{Sledgehammer} component of the Isabelle/HOL theorem prover.

Despite the idea's promise, we cannot simply invoke existing relevance filter
algorithms in our theory exploration setting. The reason is that relevance
filtering is a supervised learning method, i.e. it would require a distinguished
expression to compare everything against. Theory exploration does not have such
a distinguished expression; instead, we are interested in relationships between
\emph{any} terms generated from a signature, and hence we must consider the
relevance of \emph{all terms} to \emph{all other terms}.

A natural fit for this task is \emph{clustering}, which attempts to group
similar inputs together in an unsupervised way. Based on their success in
discovering relationships and patterns between expressions in Coq and ACL2 (in
the ML4PG and ACL2(ml) tools respectively), we hypothesise that clustering
methods can fulfil the role of relevance filters for theory exploration:
intelligently breaking up large signatures into smaller ones more amenable to
brute force enumeration, such that related expressions are explored together.
\fi

### 2019-11-29 PhDThesis/litreview_automatedreasoning.tex
A bunch of cleanup

The successor of AM is Eurisko, a discovery system intended to be useful in more
general domains than just mathematical theory formation. Despite claiming some
early successes % FIXME Traveller TCS
work on Eurisko was put on hold due to a lack of machine-readable data about
real world systems for it to work with % FIXME Find something quotable
. This lead to the Cyc project, in an attempt to encode such knowledge in a
semantically rich and consistent form. Cyc is an ongoing project, with parts of
its database made available via OpenCyc.

>>>>

The successor of AM is Eurisko~\cite{lenat83}, a discovery system intended to be
useful in more general domains than just mathematical theory formation. Despite
claiming some early successes (such as winning the ``Traveller TCS'' game) work
on Eurisko was put on hold in favour of the Cyc project~\cite{lenat1995cyc}: an
attempt to encode real-world knowledge in a semantically rich and consistent
form, for consumption by Eurisko-like algorithms. Cyc is an ongoing project,
with parts of its database made available via OpenCyc.

### 2019-11-29 PhDThesis/related_formalstats.tex
A bunch of cleanup

\iffalse
\subsection{Clustering}
\label{sec:clustering}

% TODO: ML4PG
\cite{journals/corr/abs-1212-3618}
% TODO: ACL2(ml) (also examples section)
\cite{heras2013proof}
\fi

>>>>

\subsection{Clustering}
\label{sec:clustering}

The most relevant related work for our recurrent clustering approach is
\mlforpg{}~\cite{journals/corr/abs-1212-3618}. This clusters expressions in the
Coq proof assistant, for presenting to the user (e.g. if they are stuck on a
proof, it may be useful to read ``similar'' proofs). Like our approach, this
transforms syntax trees into a matrix of values, which is then flattened into a
vector. We use the same idea of clustering referred-to definitions in order to
attain transitive similarity; however, our approach must also handle mutual
recursion (which is much more restricted in Coq, and hence unsupported by
\mlforpg{}).

The other prior work for recurrent clustering is\aclml{}~\cite{heras2013proof},
which is similar to \mlforpg{} but applies to ACL2 (Common Lisp) and is hence
untyped. The algorithm also differs, since it \emph{summarises} trees, rather
than directly encoding them to vector elements.
